### 2020-01-16 ISSTA/experiments.tex
more

Table \ref{tab:overviewexp} shows a summary of our experimental
results.  In all cases, our approach was able to detect interesting
nondeterminism (in fact, all nondeterminism of which we are aware).  The overhead on testing imposed by nondeterminism
checks varied considerably, but in the worst case did not double the
average cost of each action (the expected overhead, since we
essentially execute each test twice), and thus never halved the amount of
testing performed.  Test reduction always reduced the size of
generated tests by 85\% or more, in less than 15 minutes, thus showing that reduction
has a large payoff at an acceptable price.

@@

Our first subject is a simple implementation of a parallel merge
sort\footnote{\url{https://gist.github.com/stephenmcd/39ded69946155930c347}}.  The
merge sort and harness can be found in the TSTL GitHub repository
\cite{tstl}, under {\tt examples/parallelsorts}.  When a sorting
algorithm is implemented using concurrency, a key question to ask is
whether the ordering of elements that are equal under the comparison
operator is consistent across sorts of the same original sequence.
This is related to the question of whether a sort is a stable sort
(preserving the original relative order of equal-by-comparison elements),
but imposes a less strict requirement on the sort; a concurrent sort
might not be stable, but might still produce the same (non-stable)
ordering from the same provided sequence, every time.  This is a
particular instance of a common expectation noted above:  while a parallel
implementation of an algorithm will likely have internal
nondeterminism (due to scheduling of work), it is often expected to
behave deterministically, from the point of view of a caller.

This experimental subject serves two purposes:  first, to show that the
nondeterminism detection features of TSTL can be used to, with
acceptable overhead, show that a parallel sort implementation provides
a consistent, deterministic ordering, and second to show that the
mechanism can also detect the nondeterminism in a parallel sort that
does exhibit nondeterministic behavior.  For the second purpose, we
implemented a very simple swap-based parallel sorting algorithm that
repeatedly has multiple threads scan through a sequence and swap
out-of-order neighbors, until the sequence is sorted (a kind of
parallel bubble-sort with a simpler termination criteria).

\begin{figure}
{\scriptsize
\begin{code}
@import mergesort
@import swapsort
%\vspace{0.1in}
pool: <val> 10
pool: <data> 10
pool: <sorted> 10
%\vspace{0.1in}
<val> := <[-10..10]>
%\vspace{0.1in}
<data> := []
<data>.append((<val>,<val>))
len(<data,2>) < len(<data,1>) -> <data>.extend(<data>)
%\vspace{0.1in}
<sorted> := mergesort.merge\_sort\_parallel(<data>)
<sorted> := swapsort.swap\_sort\_parallel(<data>)
<sorted>
%\vspace{0.1in}
property: swapsort.is\_sorted(<sorted>)
\end{code}
}
\caption {Complete TSTL harness for checking two parallel sorting
  algorithms for nondeterminism.}
\label{fig:parallelcode}
\end{figure}

Figure \ref{fig:parallelcode} shows the complete TSTL harness for
checking the parallel sorts.  This harness produces sequences of
integer pairs, where only the first integer is used in the comparison
operator of the sorting algorithms, and checks that the results of
{\tt sort\_parallel} calls are in fact sorted, using the {\tt
  swapsort}'s function for checking if a sequence is sorted.

Running TSTL's random tester without checking for determinism, we can
see that both sorting algorithms always produce sorted output.  If we run with {\tt --checkDeterminism --determinismDelay 0}, however, a
test case showing the {\tt swapsort} behaving nondeterministically is
almost instantly produced (it takes less than 3 seconds), thus
answering yes to {\bf  RQ1}.  If we comment out the call to {\tt
  swap\_sort\_parallel} and only check the parallel merge sort, however, TSTL
finds no nondeterministic behavior.  Running tests of the merge sort
with nondeterminism detection turned on results in a slowdown of approximately
40\% ({\bf RQ2}.

Reducing a lengthy test showing nondeterminism of the swapsort to a
short test case using delta-debugging usually takes less than two
minutes.  If the number of attempts to produce nondeterminism is
increased in the reducer using the {\tt --determinismTries 10} option,
the resulting test will be usually be very compact, in this case
reducing the original test from 56 to 8 steps.

@@

For this simple example, more sophisticated delta-debugging strategies
are not needed; we only want to see the code behaving
nondeterministically, and do not care about preserving a higher probability of nondeterminism.

@@

Using TSTL's harness for {\tt redis-py}, both {\tt redis-py} and Redis
itself can be tested; unfortunately, generating stand-alone
high-coverage regression tests for {\tt redis-py} and Redis using TSTL
has proven difficult, as numerous Redis commands introduce nondeterministic
behavior:  thus the resulting tests are very often \emph{flaky}.   Using {\tt tstl\_afl\_fuzz}, we can see that {\tt redis-py} has a 
stability of only 56.26\%, a clear indicator of significant nondeterminism. Some
of the problematic commands are obvious simply by inspection (e.g.,
{\tt randomkey, srandmember}).  And, given an understanding of Redis
semantics, it is also clear that the various
commands producing data with a limited lifetime (e.g., {\tt expire,
  pexpire}) introduce timing-based nondeterminism.

@@

Using the original TSTL harness, just over 20\% of all {\tt redis-py}
regression tests (of length 200) generated were flaky.  Using the
nondeterminism checker to reduce flaky tests to a minimal set allows
us to simply set up an overnight run that will, in 12 hours, produce
a set of minimal nondeterministic tests exposing the root sources of
nondeterminism/flakiness: we set the harness to repeatedly run, then
produce a regression test.  If the test is flaky (exhibits
nondeterminism in result, over a set of 3 runs), we use the TSTL reducer to shrink it to a
normaliized \cite{onetest} test.  At the end of the testing period,
all distinct tests exhibiting nondeterminism are reported to the
user.  The process reduces the large number of different tests
exhibiting flakiness to a small set of tests, each short, showing a
different cause of nondeterminism.  The set produced is not perfect,
there is some duplication of causes, but it is sufficient for
identifying causes of nondeterminism.

This process is unlikely to truly require 12
hours, of course; our point is that it is completely automated and
can be given time proportional to the need to find even
low-probability sources of nondeterminism.  The actual amount of time given
may be smaller, or, for that matter, larger (leaving an important
library to self-document nondeterminism over a weekend is hardly a bad
practice).  In this case, the probability of flakiness
is not important, just the possibility, so we let the reducer
create a minimal test with some non-zero (but possibly very small)
observed probability of nondeterminism.  Our 12 hour run identified
all sources of nondeterminism successfully, according to a series of
48 hour runs performed on the version of the harness we modified to
remove nondeterminism, which produced no flaky tests.  We determined that the mean time to find all
the sources of nondeterminism and produce minimal tests 3.2
hours, which is not trivial, but also, for this purpose, not
excessive.  Our answer to {\bf RQ1} is thus a strong affirmative.

@@

We removed 11 calls from the {\tt redis} harness, and  modified 2
calls, after
identification of sources of nondeterminism.  After making these changes, no flakiness was observed in a sample of over 2,000 length
200 tests.   Moreover, because the removals were limited to actually
observed sources of flakiness, the code coverage loss was minimal.
Mean branch coverage, for regression tests of length 200, was reduced
by less than 5 branches (a less than 1\% decrease).  Obviously, it is
impossible to test the removed calls, but the overall coverage loss is
both minimal and known, and can be made up for using specially-crafted
tests (for instance, wrapping the problematic calls in a way that does
not check the values, or adding a delay after an expiration to allow
the data to expire).  As a price to pay for the ability to produce
fast-executing high-coverage full regression tests (not just tests for
crashes and unexpected exceptions), this seems acceptable.  AFL stability was 56.26\% with the harness allowing 
nondeterminism, but rose to 98.32\% using the harness with 
nondeterministic operations removed.

@@

As to {\bf RQ3}, Figures \ref{fig:r1}-\ref{fig:r3} show delta-debugging of a typical
{\tt redis-py} nondeterministic test, originally with 287 steps.  The initial test behaves
nondeterministically about half the time.  Figure \ref{fig:r1} shows
that this is a non-monotonic reduction problem, where removing steps
can either increase or decrease the probability of nondeterminism.  At
first, unmodified delta-debugging actually improves the probability of
failure, but eventually it produces a test of length 16 that only behaves
nondeterministically 24\% of the time.  The reduction takes only 38
seconds.  For the purpose of identifying commands leading to
nondeterminism, this is acceptable.  However, if we were actually
debugging a complex nondeterminism bug in Redis itself, we might want
a more reliably nondeterministic test.  Using the same parameters as
in Section \ref{sec:pubbias}, we see the same pattern.  Simply making
a predicate that ``forces'' the probability to remain high, with a
large number of samples, does not work (Figure \ref{fig:r3}) and
requires over 2000 seconds to produce a test with an even worse
probability of nondeterminism.  Using 10 samples and 10 replications,
on the other hand, actually improves the probability of
nondeterministic behavior, and (in this case) even produces a smaller test (only 14 steps) in just
over 10 minutes.

@@

These results, of course, depend on the parameters of the check for
nondeterminism.  If we increase the delay to 0.01 seconds and the
number of replays of a test used to check for nondeterminism, we
almost double the number of nondeterministic tests discovered.  Both
visible value and final state approaches find 40 nondeterministic
tests, though the overlap of tests thus detected is not quite perfect
--- each method detects two tests the other does not.  The difference
in overhead, interestingly, is very similar:  0.48\%; however, under
these parameters, visible value nondeterminism is actually cheaper on average
than final state nondeterminism, due to early termination of the
additional replays, when nondeterminism is detected (the difference is
again significant by Wilcoxon test, with $p < 1.44 \times 10^{-10}$).

@@

\begin{figure}[t]
{\scriptsize
\begin{code}
@import inference\_algs
@import datarray
%\vspace{0.1in}
<@
def flatten\_and\_sort(v):
    return (sorted(map(flatten\_and\_sort,v),key=repr) if type(v) in [list,tuple] else
                (flatten\_and\_sort(list(v.items())) if type(v) == dict else v))

def psplit(P):
    return ([P,1.0-P])
@>
%\vspace{0.1in}
pool: <P> 3
pool: <cpts> 3
pool: <evidence> 3 OPAQUE
pool: <ename> 3
pool: <event> 3
%\vspace{0.1in}
<P> := 0.01 * <[0..100]>
<ename> := "E" + str(<[1..5]>)
\{Exception\} <event> := datarray.DataArray(psplit(<P>), axes = [<ename>])
\{Exception\} <ename,1>!=<ename,2> -> <event> := [datarray.DataArray([[psplit(<P>)],
   psplit(<P>)],[<ename>,<ename>])]
<cpts> := []
~<cpts>.append(<event>)
<evidence> := \{\}
~<evidence>.update([(<ename>,0)])
%\vspace{0.1in}
\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_simple(<cpts>,
   <evidence>)))
\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_sumproduct(<cpts>,
   <evidence>)))
\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_jtree(<cpts>,
   <evidence>)))
\end{code}
}
\caption {Complete TSTL harness for finding the hash-order bug in the datarray
  inference algorithms.}
\label{hashbug}
\end{figure}

@@

Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
  datarray} algorithms (it can be found in the {\tt
  examples/datarray\_inference} directory in the TSTL repository \cite{tstl}).  The {\tt flatten\_and\_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.  Running this harness using TSTL's {\tt
  --checkProcessDeterminism} flag required less than 10 seconds on
average to produce a test exhibiting process-level nondeterminism in
the {\tt calc\_marginals\_sumproduct} function (the only broken
algorithm), answering {\bf RQ1} again in the affirmative.  Reducing this 60 step test to a minimal test of only 6 steps,
showing an extremely simple input producing the issue, 
required another 92 seconds ({\bf RQ3}).  Interestingly, the way that {\tt
  python-afl} is used in TSTL means that AFL stability is 100\% even
for the original, broken, code,
because the {\tt PYTHONHASHSEED} has already been chosen before the
fork in AFL executions.

Removing the nondeterministic call, we can see that the cost of
checking for process nondeterminism, with no delay between operations,
is high, a 93\% slowdown({\bf RQ2}.  This is due to the high cost of subprocess
creation and communication. However, even with this essentially worst
case behavior (where test runtime is very low compared to process overheads),
we sacrifice less than half of the test executions for a given time budget.

>>>>

Table \ref{tab:overviewexp} shows a summary of our experimental
results.  In all cases, our approach was able to detect interesting
nondeterminism (in fact, all nondeterminism of which we are aware).  The overhead on testing imposed by nondeterminism
checks varied considerably, but in the worst case did not double the
average cost of each action (the expected overhead, since we
essentially execute each test twice), and thus never halved the amount of
testing performed.  Test reduction always reduced the size of
generated tests by 85\% or more, in less than 15 minutes, thus showing that reduction
has a large payoff at an acceptable price.  Values here and discussed
below are, unless otherwise noted, the mean of 30 runs; we omit variance due to little
variance exhibited, but can provide full details in our replication package.

@@

Our first subject is an implementation of a parallel merge
sort\footnote{\url{https://gist.github.com/stephenmcd/39ded69946155930c347}}. When a sorting
algorithm is implemented using concurrency, a key question to ask is
whether the ordering of (equal under comparison) elements is consistent.
This is related to the question of whether a sort is stable,
but imposes a less strict requirement; a concurrent sort
might not be stable, but might still produce the same (non-stable)
ordering from the same provided sequence, every time.  This is a
particular instance of a common expectation:  while a parallel
implementation of an algorithm will likely have internal
nondeterminism, it is often expected to
behave deterministically, from the point of view of a caller.

This experimental subject serves two purposes:  first, to show that the
nondeterminism detection features of TSTL can be used to, with
acceptable overhead, show that a parallel sort implementation provides
a consistent, deterministic ordering, and second to show that the
mechanism can also detect the nondeterminism in a parallel sort that
does exhibit nondeterministic behavior.  For the second purpose, we
implemented a very simple swap-based parallel sorting algorithm that
repeatedly has multiple threads scan through a sequence and swap
out-of-order neighbors, until the sequence is sorted (a kind of
parallel bubble-sort).

\begin{figure}
{\scriptsize
\begin{code}
@import mergesort
@import swapsort

pool: <val> 10
pool: <data> 10
pool: <sorted> 10

<val> := <[-10..10]>

<data> := []
<data>.append((<val>,<val>))
len(<data,2>) < len(<data,1>) -> <data>.extend(<data>)

<sorted> := mergesort.merge\_sort\_parallel(<data>)
<sorted> := swapsort.swap\_sort\_parallel(<data>)
<sorted>

property: swapsort.is\_sorted(<sorted>)
\end{code}
}
\caption {Complete TSTL harness for checking two parallel sorting
  algorithms for nondeterminism.}
\label{fig:parallelcode}
\end{figure}

Figure \ref{fig:parallelcode} shows the complete TSTL harness for
checking the parallel sorts.  This harness produces sequences of
integer pairs, where only the first integer is used in the comparison
operator of the sorting algorithms, and checks that the results of
{\tt sort\_parallel} calls are sorted..
Running TSTL's random tester without checking for determinism, we can
see that both sorting algorithms always produce sorted output.  The
horizontal determinism detector, however, always generates a 
test case showing the {\tt swapsort} behaving nondeterministically in
less than 3 seconds, thus
answering yes to {\bf  RQ1}.  If we comment out the call to {\tt
  swap\_sort\_parallel} and only check the parallel merge sort, however, TSTL
finds no nondeterministic behavior.  Running tests of the merge sort
with nondeterminism detection turned on results in a slowdown of approximately
40\% ({\bf RQ2} with little variance.

Reducing a lengthy test showing nondeterminism of the swapsort usually takes less than two
minutes (out of 10 trials, only one took four minutes).  If the number of attempts to produce nondeterminism is
increased in the reducer, the resulting test will be usually be very compact, e.g.
reducing the original test from 56 to 8 steps.

@@

%For this simple example, more sophisticated reduction strategies
%are not needed; we only want to see the code behaving
%nondeterministically, and do not care about preserving a higher probability of nondeterminism.

@@

Using TSTL's harness for {\tt redis-py}, both {\tt redis-py} and Redis
itself can be tested; unfortunately, generating stand-alone
high-coverage regression tests for {\tt redis-py} and Redis using TSTL
has proven difficult, as numerous Redis commands introduce nondeterministic
behavior:  thus the resulting tests are very often \emph{flaky}.
Using AFL, we can see that {\tt redis-py} has a 
stability of only 56.26\%, a clear indicator of significant nondeterminism. Some
of the problematic commands are obvious simply by inspection (e.g.,
{\tt randomkey, srandmember}).  And, given an understanding of Redis
semantics, it is also clear that the various
commands producing data with a limited lifetime (e.g., {\tt expire,
  pexpire}) introduce timing-based nondeterminism.

@@

Using the original TSTL harness, just over 20\% of all {\tt redis-py}
regression tests (of length 200) generated were flaky.  Using the
nondeterminism checker to reduce flaky tests to a minimal set allows
us to simply set up an overnight run that will, in 12 hours, produce
a set of minimal nondeterministic tests exposing the root sources of
nondeterminism/flakiness: we set the harness to repeatedly run, then
produce a regression test.  If the test is flaky (exhibits
nondeterminism in result, over a set of 3 runs), we use the TSTL reducer to shrink it to a
normaliized \cite{onetest} test.  At the end of the testing period,
all distinct tests exhibiting nondeterminism are reported to the
user.  The process reduces the large number of different tests
exhibiting flakiness to a small set of tests, each short, showing a
different cause of nondeterminism.  The set produced is not perfect:
There is some duplication of causes, but it is sufficient for
identifying causes of nondeterminism.

This process is unlikely to truly require 12
hours, of course; our point is that it is completely automated and
can be given time proportional to the need to find even
low-probability sources of nondeterminism.  The actual amount of time given
may be smaller, or, for that matter, larger (leaving an important
library to self-document nondeterminism over a weekend is hardly a bad
practice).  In this case, the probability of flakiness
is not important, just the possibility, so we let the reducer
create a minimal test with some non-zero (but possibly very small)
observed probability of nondeterminism.  Our 12 hour run identified
all sources of nondeterminism successfully, according to a series of
48 hour runs performed on the version of the harness we modified to
remove nondeterminism, which produced no flaky tests.  We determined that the mean time to find all
the sources of nondeterminism and produce minimal tests is 3.2
hours, which is not trivial, but also, for this purpose, not
excessive.  Our answer to {\bf RQ1} is thus a strong affirmative.

@@

We removed 11 calls from the {\tt redis-py} harness, and  modified 2
calls, after
identification of sources of nondeterminism.  After making these changes, no flakiness was observed in a sample of over 2,000 length
200 tests.   Moreover, because the removals were limited to actually
observed sources of flakiness, the code coverage loss was minimal.
Mean branch coverage, for regression tests of length 200, was reduced
by less than 5 branches (a less than 1\% decrease).  Obviously, it is
impossible to test the removed calls, but the overall coverage loss is
both minimal and known, and can be made up for using specially-crafted
tests (for instance, wrapping the problematic calls in a way that does
not check the values, or adding a delay after an expiration to allow
the data to expire).  As a price to pay for the ability to produce
fast-executing high-coverage full regression tests (not just tests for
crashes and unexpected exceptions), this seems acceptable.  AFL stability was 56.26\% with the harness allowing 
nondeterminism, but rose to 98.32\% using the harness with 
nondeterministic operations removed.

@@

As to {\bf RQ3}, Figures \ref{fig:r1}-\ref{fig:r3} show delta-debugging of a typical
{\tt redis-py} nondeterministic test, originally with 287 steps.  The initial test behaves
nondeterministically about half the time.  Figure \ref{fig:r1} shows
that this is a non-monotonic reduction problem, where removing steps
can either increase or decrease the probability of nondeterminism.  At
first, unmodified delta-debugging actually improves the probability of
failure, but eventually it produces a test of length 16 that only behaves
nondeterministically 24\% of the time.  The reduction takes only 38
seconds.  For the purpose of identifying commands leading to
nondeterminism, this is acceptable.  However, if we were actually
debugging a complex nondeterminism bug in Redis itself, we might want
a more reliably nondeterministic test.  Using the same parameters as
in Section \ref{sec:pubbias}, we see the same pattern.  Simply making
a predicate that ``forces'' the probability to remain high, with a
large number of samples, does not work (Figure \ref{fig:r3}) and
requires over 2,000 seconds to produce a test with an even worse
probability of nondeterminism.  Using 10 samples and 10 replications,
on the other hand, actually improves the probability of
nondeterministic behavior, and (in this case) even produces a smaller test (only 14 steps) in just
over 10 minutes.

@@

These results, of course, depend on the parameters of the check for
nondeterminism.  If we increase the delay to 0.01 seconds and the
number of replays of a test used to check for nondeterminism, we
almost double the number of nondeterministic tests discovered.  Both
visible value and final state approaches find 40 nondeterministic
tests, though the overlap of tests thus detected is not quite perfect---each method detects two tests the other does not.  The difference
in overhead, interestingly, is very similar:  0.48\%; however, under
these parameters, visible value nondeterminism is actually cheaper on average
than final state nondeterminism, due to early termination of the
additional replays, when nondeterminism is detected (the difference is
again significant by Wilcoxon test, with $p < 1.44 \times 10^{-10}$).

@@

\begin{figure}[t]
{\scriptsize
\begin{code}
@import inference\_algs
@import datarray
<@
def flat\_sort(v):
    return (sorted(map(flat\_sort,v),key=repr) if type(v) in [list,tuple]
                else (flat\_sort(list(v.items())) if type(v) == dict else v))

def psplit(P):
    return ([P,1.0-P])
@>
pool: <P> 3
pool: <cpts> 3
pool: <evidence> 3 OPAQUE
pool: <ename> 3
pool: <event> 3

<P> := 0.01 * <[0..100]>
<ename> := "E" + str(<[1..5]>)
\{Exception\} <event> := datarray.DataArray(psplit(<P>),axes=[<ename>])
\{Exception\} <ename,1>!=<ename,2> -> <event> := [datarray.DataArray([[
   psplit(<P>)], psplit(<P>)],[<ename>,<ename>])]
<cpts> := []
~<cpts>.append(<event>)
<evidence> := \{\}
~<evidence>.update([(<ename>,0)])

\{Exception\} print(flat\_sort(inference\_algs.calc\_marginals\_simple(<cpts>,
   <evidence>)))
\{Exception\} print(flat\_sort(inference\_algs.calc\_marginals\_sumproduct(<cpts>,
   <evidence>)))
\{Exception\} print(flat\_sort(inference\_algs.calc\_marginals\_jtree(<cpts>,
   <evidence>)))
\end{code}
}
\caption {Complete TSTL harness for finding the hash-order bug in the datarray
  inference algorithms.}
\label{hashbug}
\end{figure}

@@

Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
  datarray} algorithms.  The {\tt flat\_sort} function is
needed because we care about actual differences in probability values,
not simply the order of list, tuple, or dictionary items.  Running
this harness consistently requires less than 10 seconds to produce a test exhibiting process-level nondeterminism in
the {\tt calc\_marginals\_sumproduct} function (the only broken
algorithm), answering {\bf RQ1} again in the affirmative.  Reducing this 60 step test to a minimal test of only 6 steps,
showing an extremely simple input producing the issue, 
required another 92 seconds ({\bf RQ3}) on average.  Interestingly, the way that {\tt
  python-afl} is used in TSTL means that AFL stability is 100\% even
for the original, broken, code,
because the {\tt PYTHONHASHSEED} has already been chosen before the
fork in AFL.  Removing the nondeterministic call, we can see that the cost of
checking for process nondeterminism, with no delay between operations,
is high, a mean 93\% slowdown ({\bf RQ2}.  This is due to the high cost of subprocess
creation and communication. However, even with this essentially worst
case behavior (where test runtime is very low compared to process overheads),
we sacrifice less than half of the test executions.
